### 2016-06-25 store-consistency.tex
edits to sec. 5

The abstract machine
of Fig.~\ref{fig:txnimp} allows operations to witness arbitrary
subsets of the global state, thus assuming the semantics of an
eventually consistent ({\sc ec}) data store\footnote{{\sc ec}
guarantees that in the absence of further updates, all reads witness
the same global state \emph{eventually}. In any finite trace, however,
there are no guarantees on what a read witnesses.}. There are however
stores, such as relational databases, that provide stronger
consistency guarantees than {\sc ec}. Like the isolation levels of
transactions, the consistency level of the underlying store also
affects the semantics of a program in non-trivial ways. In this
section, we demonstrate how the semantics of stronger stores with
on-demand weak isolation can be captured in our operational model
and reasoning framework.

@@

Firstly, we observe that the semantics of stronger stores can be
captured by store-specific consistency constraints, along with
transaction-specific isolation constraints, via the trace invariant
$\I$. In particular, we split $\I$ into two components: (1).  $\I_s$,
the store-specific invariant, and (2). $\I_c$, the program-specific
(or, client-specific) invariant, to capture consistency and isolation
constraints, respectively. $\I$ is now a conjunction:
  $\I \;=\; \lambda\E.~\I_s(\E) \wedge \I_c(\E)$.
While the program-specific trace invariant ($\I_c$) and remains an
invariant regardless of the store, the store specific invariant
($\I_s$) changes from store to store depending on the consisteny
level. $\I_s$ for an {\sc ec} store is $true$. Stronger stores, such
as an {\sc sc} store, have non-trivial $\I_s$.

@@

In practice, relational databases resolve the conflict by prioritizing
weak isolation (thus, concurrency and performance) over strong
consistency, so the execution traces do not necessarily satisfy {\sc
sc}. In particular, visibility constraints imposed by {\sc sc} are
violated if (and only if) they are found to be in conflict with the
constraints imposed by the isolation level. In the context of
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term
\emph{recency}~\cite{bailisvldb} is often used in place of consistency
to capture the best-effort nature of the {\sc sc} guarantee on
databases.}. We generalize this approach to any $\I_s$ and $\I_c$ by
defining a \emph{maximum visiblity principle} to determine an
acceptable weakening of $\I_s$ in case of a conflict with $\I_c$.  The
principle requires the weakened consistency guarantee ($\I_s'$) of the
store to enforce all visibility relationships imposed by the actual
consistency guarantee ($\I_s$), unless enforcing such a relationship
violates $\I_c$. Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}

Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes a practical {\sc sc} store, such as a relational database.

@@

In this section, we present an intuitive explanation of the soundness
result in context of eventually consistent replication. 
% In other words, we explain why our approach is a sound way of
% reasoning about transactions under eventually consistent replication
% despite our system model not including the standard artifacts
% associated with a replicated store.

>>>>

The abstract machine of Fig.~\ref{fig:txnimp} allows operations to
witness arbitrary subsets of the global state, thus assuming the
semantics of an eventually consistent ({\sc ec}) data
store\footnote{{\sc ec} guarantees that in the absence of further
updates, all reads witness the same global state \emph{eventually}. In
any finite trace, however, there are no guarantees on what a read
witnesses.}. There are however stores, such as relational databases,
that provide stronger consistency guarantees than {\sc ec}. Like the
isolation levels of transactions, the consistency level of the
underlying store also affects the semantics of a program in
non-trivial ways. In this section, we demonstrate how the semantics of
stronger stores with on-demand weak isolation can be captured in our
operational model so that we can use the reasoning framework developed
in \S\ref{sec:reasoning} to reason about such stores.

@@

Firstly, we observe that the semantics of stronger stores can be
captured by store-specific consistency constraints, along with
transaction-specific isolation constraints, via the trace invariant
$\I$. In particular, we split $\I$ into two components: (1).  $\I_s$,
the store-specific invariant, and (2). $\I_c$, the program-specific
(or, client-specific) invariant, to capture consistency and isolation
constraints, respectively. $\I$ is now a conjunction:
  $\I \,=\, \lambda\E.~\I_s(\E) \wedge \I_c(\E)$ (often simplified to
  $\I \,=\, \I_s\wedge\I_c$).
While the program-specific trace invariant ($\I_c$) and remains an
invariant regardless of the store, the store specific invariant
($\I_s$) changes from store to store depending on the consisteny
level. $\I_s$ for an {\sc ec} store is $true$. Stronger stores, such
as an {\sc sc} store, have non-trivial $\I_s$.

@@

In practice, relational databases resolve the conflict by prioritizing
weak isolation (thus, concurrency and performance) over strong
consistency, so the execution traces do not necessarily satisfy {\sc
sc}. In particular, visibility constraints imposed by {\sc sc} are
violated if (and only if) they are found to be in conflict with the
constraints imposed by the isolation level. In the context of
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store needs to make $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term
\emph{recency}~\cite{bailisvldb} is often used in place of consistency
to capture the best-effort nature of the {\sc sc} guarantee on
databases.}. We generalize this approach to any $\I_s$ and $\I_c$ by
defining a \emph{maximum visiblity principle} to determine an
acceptable weakening of $\I_s$ in case of a conflict with $\I_c$.  The
principle requires the weakened consistency guarantee ($\I_s'$) of the
store to enforce all visibility relationships imposed by the actual
consistency guarantee ($\I_s$), unless enforcing such a relationship
violates $\I_c$. The formal definition of the principle is relegated
to the supplementary in the interest of space, but it can be
understood in the context of an {\sc sc} store, where it weakens the
store invariant to the following:
% Formally:
% \begin{definition}
% $\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
% weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
% \begin{itemize}
%   \item $\I_s'$ is weaker than $\I_s$: 
%       $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
%   \item In every trace $\E$ that satisfies $\I_c$, and for every pair
%   of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
%   $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
%   extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
%   $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
%   on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
%   (considering that the machine never violates them)? Encode the
%   specifications in Z3 and make sure they are consistent.}}:
%   \begin{smathpar}
%   \begin{array}{l}
%   \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
%     \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
%     \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
%     \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
%   \end{array}
%   \end{smathpar}
% \end{itemize}
% \end{definition}

% Applying this principle, we can weaken {\sc sc} to obtain the
% following store trace invariant ($\I_s$) for an {\sc sc} store whose
% isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes relational database-like stores.

@@

In this section, we present an intuitive explanation of the soundness
result in context of eventually consistent replication.
